---
title: "Simulating and fitting LRCQ and LRCP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating and fitting LRCQ and LRCP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(2026)
```

```{r}
library(lrcpq)
```

## The model

For `q` GWAS of sample sizes `n_a` on `m` SNPs with LD matrix `R`, the
genetic effects follow a matrix normal distribution

    B ~ MN(0, U = W^1/2 Rβ W^1/2, V = H_g / M)

where `W = diag(w)` holds the per-SNP phenome-wide enrichment (mean 1),
`Rβ` the correlation of genetic effects between SNPs, `H_g` the genetic
covariance between traits (diagonal = h²) and `M` the number of SNPs the
heritability is spread over. The Z-scores are

    Z = R B diag(√n) + E,   E ~ MN(0, R, C)

with `C` the inter-GWAS intercepts (sample overlap times phenotypic
correlation). This gives the moment equation the estimators regress on:

    E[z_a z_b'] = c_ab R W^1/2 Rβ W^1/2 R + C_ab R,   c_ab = √(n_a n_b) h_ab / M

LRCQ estimates `w` from the squared Z-scores (`Rβ = I`); LRCP estimates
off-diagonal entries of `Rβ` between distal SNPs from Z cross-products.

## Simulating data

```{r}
blocks <- make_ld_ar1(rep(50, 20), rho = 0.5)   # 20 toy LD blocks of 50 SNPs
ld     <- ld_from_blocks(blocks)
w      <- make_w(ld$m, prop_nonzero = 0.1, dist = "lnorm")
q      <- 50
gc     <- make_gcov(q, type = "cluster", rg = 0.3)   # h2 ~ Beta, two trait clusters
C      <- make_intercept(q, overlap = 1, rp = 0.1)    # UK Biobank-like full overlap
sim    <- simulate_lrcpq(blocks, n = rep(3e5, q), w = w, gcov = gc$gcov,
                         intercept = C, M = 5e4)
dim(sim$Z)
```

`M = 5e4` treats this 1,000-SNP region as 2% of a 50,000-SNP genome, which
keeps per-SNP signal at a realistic scale. Using `M = m` for a small region
would put a whole trait's heritability on a few SNPs.

## Stages 1 and 2

In real data the genetic covariance and intercepts come from genome-wide
LD score regression, `ldsc_matrix()`, or from published estimates such as
the Pan-UKB h² manifest. Here we use the true values.

## Stage 3: LRCQ

```{r}
fit <- lrcq(sim$Z, ld, sim$n, gc$gcov, M = 5e4, intercept = C,
            method = "wls", rectify = "C")
head(fit)
cor(fit$w_raw, w)
mean((fit$w_raw - w) / fit$se)       # standardised error, about 0
sd((fit$w_raw - w) / fit$se)         # about 1 when SEs are calibrated
```

```{r, fig.width = 5, fig.height = 4}
plot(w, fit$w_raw, pch = 16, cex = 0.5, xlab = "true w", ylab = "estimated w")
abline(0, 1, col = "red")
```

Options:

- `method`: `"ols"`, `"wls"` (one reweighting, default), `"irls"`, or
  `"equal"` (the simple mean of trait-specific enrichments; OLS and WLS
  target the power-weighted mean).
- `rectify`: methods A, B or C for negative estimates (`rectify_w()`);
  `w_raw` is always returned.
- `tag_r2`: estimate per clump of SNPs in tight LD (`r² ≥ tag_r2`) instead of
  per SNP, which is the identifiable target when LD is near 1.
- `cross_terms = TRUE, Rb = ...`: add the `Rβ` cross terms of chapter 6.

## Stage 4: LRCP between distal SNPs

Two regions with three SNP pairs whose effects correlate at 0.6, -0.4 and 0:

```{r}
b2 <- make_ld_ar1(c(30, 30), 0.5)
w2 <- numeric(60); i1 <- c(5, 12, 20); i2 <- 30 + c(8, 15, 25); w2[c(i1, i2)] <- 10
Rb <- as.matrix(make_Rb(60, "pairs", pairs = cbind(i1, i2), pair_rho = c(0.6, -0.4, 0)))
g2 <- make_gcov(30, h2 = rep(0.3, 30))$gcov
est <- t(replicate(40, {
  s <- simulate_lrcpq(b2, n = rep(3e5, 30), w = w2, Rb = Rb, gcov = g2, M = 2e4)
  f <- lrcp_distal(s$Z[1:30, ], s$Z[31:60, ], b2[[1]], b2[[2]],
                   w2[1:30], w2[31:60], rep(3e5, 30), g2, 2e4)
  c(diag(f$rho), diag(f$se))
}))
rbind(mean = colMeans(est[, 1:3]), sd = apply(est[, 1:3], 2, sd),
      mean_se = colMeans(est[, 4:6]))
```

`lrcp_distal()` offers GLS (default), OLS and WLS; `lrcp_mle()` fits the
Gaussian likelihood for a window pair with a positive semi-definite
parameterisation of `Rβ` and is about twice as efficient. `pair_sum =
"chapter"` reproduces the thesis equations, which count each SNP pair twice
and return half the correlation.

## Whole-region pipelines

`run_lrcq()` wraps stages 1 to 3 (estimating stage 1-2 with `ldsc_matrix()`
when `gcov` is not given), and `run_lrcp()` screens the stage-3 output on
`w` and its z-score, then fits every pair of candidate blocks that share no
LD, with Benjamini-Hochberg and Benjamini-Yekutieli q-values.

## Power

```{r}
power_lrcq(w = c(2, 5, 10), n = 3e5, h2 = 0.2, qe = 100)
power_lrcp(rho = c(0.2, 0.5), w = 10, n = 3e5, h2 = 0.2, qe = 100)
```
